Methods/Lit Review

### 2020-01-27 chapters/03_methods.Rmd
update methods, big picture mostly complete

---
title: "Methods/Lit Review"
author: "Raven McKnight"
output:
  word_document:
    toc: yes
  pdf_document: null
  html_document:
    theme: paper
    toc: yes
    toc_float: yes
---

@@

Each model presented in this honors is built upon a simple Poisson regression. Poisson regressions are used to model count data. The model assumes that the natural log of the mean of response variable $Y$ can be modeled by a linear combination of predictors. Poisson regressions are a form of generalized linear models with the natural log as its link function. For response variable $Y$ and covariates $x_1, x_2, ..., x_k$, the basic Poisson regression can be written

@@

The most obvious assumption made in a Poisson regression is that response variable $Y$ follows a Poisson distribution. The primary assumption this makes about our data is that $\text{Var}(Y) = \text{E}(Y)$. When this assumption is not met, we can either use a generalization of the Poisson distribution, such as a Negative-Binomial distribution, or add a parameter to account for extra-Poisson variance.

For the purposes of this study, we can add a set of heterogeneous random effects $\theta$ to account for overdispersion (where Var($Y$) > E($Y$)). Following Morris et al (2019), the second line of **1** can be written

@@

where $\theta$ is given a vague normal prior centered around 0, such as $\theta \sim$ Normal(0, 1). The addition of these random effects improve fits on data with higher-than-expected variance. With sufficiently vague priors, $\theta$ can account for most or all overdispersion not accounted for by $\beta_0 + \sum_{k=1}^{K}x_k^T\beta_k$.

@@

## Spike and Slab

@@

## Horseshoe

@@

The horseshoe prior, proposed by **Carvalho et all, 2010** is a popular choice in Bayesian literature, in part because it is similar to the "gold standard" spike-and-slab. Following the notation of **Vehtari & Pironeen**, the horseshoe prior can be expressed

@@

# Spatial Structure

@@

### This whole section follows Morris et al (probably too) closely right now

We can encode spatial information into Bayesian models in several ways. Conditional Autoregressive (CAR) priors are one of the most widespread Bayesian methods for modeling spatial autocorrelation. Conditional Autoregressive models were introduced in Besag 1974 and remain perhaps the primary method for Bayesian areal data modeling. CAR models are used as $priors$, not as the actual observation model.

@@

Generally, CAR models use binary neighborhood relationships encoded in a neighborhood matrix. Areal units $i$ and $j$ are considered neighbors if they share a boundary. For strictly rectangualar lattice data, we often choose either a "Queens" or "Rooks" neighborhood structures **(include graphic)**. For irregularly shaped areal units, such as census tracts, we generally default to a queen neighborhood structure wherein areal units that share *any* points of contact are deemed neighbors. For $M$ regions represented in a $M$x$M$ neighborhood matrix $W$, $w_{ij} = 1$ if regions $i$ and $j$ are neighbors and 0 otherwise. Matrix $W$ is symmetric. Note that regions $i$ are not considered neighbors with themselves.

The spatial interactions between $i$ and $j$ are represented by random variable $\phi$. Here, $\phi$ is a vector $\phi = \phi_1, \phi_2, ..., \phi_M$. The distribution of each $\phi_i$ is determined by the sum of its neighbors values such that

# having a notation crisis

$$
\phi_i \text{ | } \phi_j, j \neq i \sim \text{Normal}(\sum_{i=1}^{M}w_{ij}\phi_j, \sigma^2) 
$$

@@

The ICAR specifies each $\phi_i$ to be normally distributed with its mean equal to the mean of its neighbors values. This is how the CAR/ICAR model "borrows strength" from geographic neighbors to smooth noisy estimates. Intuitively, the variance of each $\phi_i$ decreases as its number of neighbors $d_i$ increases. The conditional distribution for each $\phi_i$ can be written

@@

Here, the variance $\sigma$ is unknown. The conditional specification of each $\phi_i$ is helpful in building intuition about the assumptions the ICAR component makes and how it performs its "smoothing". The joint distribution can be rewritten

@@

This pairwise difference specification is also helpful in building intuition about how ICAR models perform smoothing.

@@

The specific ICAR model used in this analysis is the Besag-York-Mollie (BYM) Poisson model. The BYM model contains random effects and a spatial ICAR term to account for both spatial and non-spatial heterogeneity. Note that in **section??*, we introduced a random effect term $\theta$. Therefore, the BYM model can be though of as decomposing $\theta$ into spatial and non-spatial components.

Specifically, the BYM model is written

@@

where $\phi$ is the addition: an ICAR component.

The BYM model is appealing in its simplicity. However, the lack of hyperpriors specified for $\theta$ and $\phi$ can make the model incredibly challenging to fit. In theory, extra-Poisson variance could be explained 100% by $\theta$, 100% by $\phi$, or anywhere in between. In *practice*, we likely have limited information about where that ratio falls. Riebler et al (2016) further explain the sampling issues faced by the BYM model faced with no hyperpriors. In short, the sampler is forced to explore all possible combinations of $\theta$ and $\phi$, no matter how unlikely a particular combination my be. In the context of fitting models with Stan, this is liekly to cause uneccesarily long computation times and perhaps to yield biased posterior estimates.

There are some suggestions in the literature for setting hyperpriors for $\theta$ and $\phi$ (Besag and Mollie, 1991; Clayton and Montomoli, 1995). The existing methods, however, tend to rely on the specific data at hand which makes the selection of hyperparameters time unnecessarily time consuming.

The BYM2 model proposed by Simpson et al and further described in Riebler et al (2016) aims to solve these sampling problems. The primary difference between the BYM and BYM2 models is the addition of a mixing parameter, $\rho$. The BYM2 model rewrites the BYM as

$$
\eta_i = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k + ((\sqrt{\frac{\rho}{s}})\theta^{*} + ((\sqrt{1-\rho})\phi)\sigma
$$

@@

A relatively vague Normal(0, 1) prior is appropriate for $\sigma$. Mixing parameter $\rho$ is given a Beta(1/2, 1/2) prior. Similar to the horseshoe,

@@

The models above can all be fit using the Stan Programming Langauge (cite). Stan is probabilistic programming language used to compute joint log probability densities.

@@

Stan utilizes the Hamiltonian Monte Carlo (HMC) and specialized No U-Turn Sampler (NUTS).

>>>>

---
title: "Methods/Lit Review"
author: "Raven McKnight"
output:
  pdf_document:
    number_sections: true
  word_document:
  html_document:
    theme: paper
    toc: yes
    toc_float: yes
---

@@

Each model presented in this honors is built upon a simple Poisson regression. Poisson regressions are a form of generalized linear models with the natural log as its link function. They are generally used to model count data. The model assumes that the natural log of the mean of response variable $Y$ can be modeled by a linear combination of predictors. For response variable $Y$ and covariates $x_1, x_2, ..., x_k$, the basic Poisson regression can be written

@@

In the basic Poisson regression, we often give $\beta_0$ and $\beta$ vague priors such as Normal($0, 1$). The exposure term $E_i$ comes from the data and does not recieve a prior.

## Overdispersion

The most obvious assumption made in a Poisson regression is that response variable $Y$ follows a Poisson distribution. In addition to requiring integer counts, this assumption requires that $\text{Var}(Y) = \text{E}(Y)$. This is often not true in practice. In the case of the Metro Transit ridership data, $\text{Var}(Y) \approx 7*\text{E}(Y)$. In this case, we call the unexpectedly large variance "overdispersion" or "extra-Poisson variance."

Data with this feature can be difficult to model directly with a Poisson regression. One option is to use a generalization of the Poisson distribution, such as a Negative-Binomial distribution, as the observation model. For the Metro Transit case study, we instead modify the latent function to include a seperate term to account for overdispersion. As such, the Poisson regression above can be rewritten to include

@@

where $\theta$ is a set of heterogenous random effects. The addition of this random effects term improves fits on data with overdispersion. With sufficiently vague priors, $\theta$ can account for most or all of the overdispersion not accounted for by $\beta_0 + \sum_{k=1}^{K}x_k^T\beta_k$.

@@

## Spike and Slab -- needs graphics

@@

## Horseshoe and Regularized Horseshoe -- needs graphics

@@

The horseshoe prior, proposed by **Carvalho et al, 2010** is a popular choice in Bayesian literature, in part because it is similar to the "gold standard" spike-and-slab. Following the notation of **Vehtari & Piironen**, the horseshoe prior can be expressed

@@

# Spatial Structure -- needs many graphics

@@

### This whole section follows Morris et al (probably too) closely right now -- encorporate more from Riebler, Simpson, Besag
We can encode spatial information into Bayesian models in several ways. Conditional Autoregressive (CAR) priors are one of the most widespread Bayesian methods for modeling spatial autocorrelation. Conditional Autoregressive models were introduced in Besag 1974 and remain perhaps the primary method for Bayesian areal data modeling. CAR models are used as $priors$, not as the actual observation model.

@@

Generally, CAR models use binary neighborhood relationships encoded in a neighborhood matrix. Areal units $i$ and $j$ are considered neighbors if they share a boundary. For strictly rectangualar lattice data, we often choose either a "Queens" or "Rooks" neighborhood structures **(include graphic)**. For irregularly shaped areal units, such as census tracts, we generally default to a queen neighborhood structure wherein areal units that share *any* points of contact are deemed neighbors. For $N$ regions represented in a $N$x$N$ neighborhood matrix $W$, $w_{ij} = 1$ if regions $i$ and $j$ are neighbors and 0 otherwise. Matrix $W$ is symmetric. Note that regions $i$ are not considered neighbors with themselves.

The spatial interactions between $i$ and $j$ are represented by random variable $\phi$. Here, $\phi$ is a vector $\phi = \phi_1, \phi_2, ..., \phi_N$. The distribution of each $\phi_i$ is determined by the sum of its neighbors values such that

$$
\phi_i \text{ | } \phi_j, j \neq i \sim \text{Normal}(\sum_{i=1}^{N}w_{ij}\phi_j, \sigma^2) 
$$

@@

The ICAR prior specifies each $\phi_i$ to be normally distributed with its mean equal to the mean of its neighbor's values. This is how the CAR/ICAR model "borrows strength" from geographic neighbors to smooth noisy estimates. Intuitively, the variance of each $\phi_i$ decreases as its number of neighbors $d_i$ increases. The conditional distribution for each $\phi_i$ can be written

@@

Here, the variance $\sigma$ is unknown. The conditional specification of each $\phi_i$ is helpful in building intuition about the assumptions the ICAR component makes and how it performs its "smoothing". The ICAR prior matches our intuition about spatial autocorrelation: areal unit $i$ is similar to the areal units surrounding it, and the more areal units surround it, the more confident we feel in that similarity. The joint distribution can be rewritten

@@

This pairwise difference specification is also helpful in building intuition about how ICAR models perform smoothing. Each $(\phi_i - \phi_j)^2$ works as a penalty term: the ???

@@

The specific ICAR model used in this analysis is the Besag-York-Mollie (BYM) Poisson model. The BYM is a "classical" spatial Bayesian method (Riebler et al, 2016). It is generally used in disease mapping studies to model rare events (ie, disease occurrence) in small areal units. The BYM capitalizes on the CAR prior's ability to smooth noisy estimates and share information across geographic units.

The latent function in the BYM model contains heterogeneous random effects as well as a spatial ICAR term in order to account for both spatial and non-spatial heterogeneity. Note that in Section 1.1, we introduced a random effect term $\theta$. Therefore, the BYM model can be though of as decomposing $\theta$ into spatial and non-spatial components.

Specifically, the BYM model replaces $\eta_i$ from equation *1* with

@@

where $\phi$ is the addition: an ICAR component. Decomposing overdispersion as such is helpful in furthering our understanding of response $Y$. When we use a simple random effect $\theta$ alone, we can accomplish a good model fit. However, $\theta$ provides no additional information about $Y$ or why it varies so much. The decomposition of $\theta$ in the BYM allows us to better quantify how much variance is white noise (ordinary random effects) versus some unmeasured confounding correlated across space.

The BYM model as written above is appealing in its simplicity. However, the lack of informative hyperpriors specified for $\theta$ and $\phi$ can make the model incredibly challenging to fit. In theory, extra-Poisson variance could be explained 100% by $\theta$, 100% by $\phi$, or anywhere in between. In *practice*, we likely have limited information about where that ratio falls. Riebler et al (2016) further explain the sampling issues faced by the BYM model faced with no hyperpriors. In short, the sampler is forced to explore all possible combinations of $\theta$ and $\phi$, no matter how unlikely a particular combination my be. In the context of fitting models with Stan, this is liekly to cause uneccesarily long computation times and perhaps to yield biased posterior estimates.

There are some suggestions in the literature for setting hyperpriors for $\theta$ and $\phi$ (Besag and Mollie, 1991; Clayton and Montomoli, 1995). The existing methods, however, tend to rely on the specific data at hand which makes the selection of hyperparameters unnecessarily time consuming.

The BYM2 model proposed by Simpson et al and further described in Riebler et al (2016) aims to solve these sampling problems. The BYM2 takes a more "fully Bayesian" approach to hyperprior/hyperparameter selection. The primary difference between the BYM and BYM2 models is the addition of a mixing parameter, $\rho$. The BYM2 model rewrites the BYM as

$$
\eta_i = \beta_0 + \sum_{k=1}^{K}x_k^T\beta_k + ((\sqrt{\frac{\rho}{s}})\theta^{*} + ((\sqrt{1-\rho})\phi^{*})\sigma
$$

@@

A relatively vague Normal(0, 1) prior is appropriate for $\sigma$. Mixing parameter $\rho$ is given a Beta(1/2, 1/2) prior. Riebler et al (2016) also propose a more complex prior for $\rho$ such that ____ Similar to the horseshoe,

@@

The models above can all be fit using the Stan Programming Langauge (cite). Stan is probabilistic programming language used to compute joint log probability densities. The analysis in this honors uses Stan Version ____.

@@

Stan utilizes Hamiltonian Monte Carlo (HMC) and specialized No U-Turn Sampler (NUTS). The technicalities of Stan's sampling process is beyond the scope of this paper. However, there are a few sampling parameters which are important to understand in order to optimize model fits. Specifically, the parameters `max_treedepth` and `adapt_delta` appear in the application portion of this honors.

The `max_treedepth` argument specifies how many steps the sampler is allowed to make on a given transition. Stan's default max treedepth is 10. When the default is used, the sampler makes up to 10 steps. If it reaches 10 leapfrog steps without ___, the sampler stops for that iteration and Stan informs the user that maximum treedepth has been exceeded. When a model experiences many iterations reaching maximum treedepth, there are several considerations to make.

The `adapt_delta` argument works hand in hand with `max_treedepth`.
